---
title: " How to calculate Z scores in R"
format: html
editor: visual
---


## Cholesterol levels in men 18-24 years of aage are normally distributed with a *mean of 178* and a *SD of 40*. ^[These problems are found in the tutorial from David Longstreet's *statisticsisfun* channel, "How to calculate Z scores" (2011), <https://www.youtube.com/watch?v=Ypf6eAP9aFg>] 


### 1. In what percentile is a man with cholesterol level of 190?

To convert a Z score to a percentile, you need to use the pnorm() function in R to calculate the cumulative probability up to the Z score. This cumulative probability represents the percentile below the Z score, not above it.

* Z = Obs - mean/SD  

* Z = `(190-178)/40`
```{r} 
#| echo: false

(190-178)/40
```
* The Z score of 0.3 is the SD above the mean.    
To convert to a percentile, use `pnorm() * 100`

* Convert Z score to Probability  
* **percentile <- pnorm(Zscore)** 

```{r}
percentile <- pnorm(0.3)
percentile
```



## 2. What is the probability that a man in this age group has a cholesterol level greater than 145?  
*i.e. All of the are to the right of 145*.  
 
```{r}
z145 <- (145-178)/40

1- pnorm(z145)
```


## 3. What is the probability that a man in this age group has a cholesterol level greater than 178?


```{r}
z178 <- (178-178)/40

1-pnorm(z178)
```



## 4. What Cholesterol level corresponds to the 75th percentile? 

$$Z = (X - mean)/SD$$

You would apply the inverse cdf to the percentile ranks to convert them to quantiles, so if you want standard normals, z=Φ−1(p)
 should do what you seem to be asking for.
 
* To convert a value, let's call it X, to the 75th percentile in R when the mean is 178 and the standard deviation is 40, you need to calculate the corresponding Z score and then find the value that corresponds to that Z score in a standard normal distribution.

* The formula to calculate the Z score is:

  - Z = (X - mean) / standard deviation

* **To convert the Z score to the corresponding percentile, you can use the *qnorm()* function in R**.  

  - The *qnorm()* function calculates the quantile (value) of a standard normal distribution given a probability.
 
 

 * Z <- qnorm(percentile/100) = Z score corresponding to the 75th percentile,   
  then, `X <- Z & SD + mean` calculates the X corresponding to that Z score.

```{r}
mean <- 178
std_dev <- 40
percentile <- 75

Z <- qnorm(percentile/100)
X <- Z * std_dev + mean
X
```




## 5. What is the probability that a man in this age group has a cholesteral level between 155 and 185?

```{r}
probability1 <- pnorm((185-178)/40) - pnorm((155-178)/40)
probability1
```




## 6. If a sample of 100 men is selected, what is the probability that the mean cholesterol level is greater than 175?


$$Z = \frac{{X - \text{mean}}}{{\frac{{\text{SD}}}{{\sqrt{\text{sample size}}}}}}$$


```{r}

Z100 <- (175-178)/(40/10)

p <- 1-pnorm(Z100)
p
```




